
# Working with Tables using the Tidyverse

In this tutorial we will introduce the **data frame**, or the object type that R uses to store tables. Most of the data you will work with in R can be represented by a table (think an excel sheet), and one of the main advantages of using R is that the data frame is a powerful and intuitive interface for tabular data. In this tutorial we will use the [tidyverse](http://www.tidyverse.org/) to manipulate and summarise tabular data. The tutorial is a companion to the [Data transformation](http://r4ds.had.co.nz/transform.html) chapter in [R for Data Science](http://r4ds.had.co.nz/).

## Prerequisites

The prerequisite for this tutorial is the `tidyverse` package. If this package isn't installed, you'll have to install it using `install.packages()`.

```{r, eval = FALSE}
install.packages("tidyverse")
```

Load the packages when you're done! If there are errors, you may have not installed the above packages correctly!

```{r}
library(tidyverse)
library(readxl)
```

Finally, you will need to obtain the example data. You will need to [download the dataset as an excel sheet](data/halifax_geochem.xlsx) and put it in the "data" folder of your current project.

```{r}
# read local version to build
halifax_geochem <- read_excel("data/halifax_geochem.xlsx")
```

It's worth mentioning a little bit about what this data frame contains, since we'll be working with it for the rest of this tutorial.

TODO: describe the halifax_geochem dataset

## Viewing a Data Frame

The variable we have just created (`halifax_geochem`) is a **tibble**, which is a table of values much like you would find in a spreadsheet (you will notice that we loaded it directly from an Excel spreadhseet). Each column in the tibble represents a variable (in this case, the core identifier, depth of the sample, age represented by that sample, and several geochemical parameters), and each row in the table represents an observation (in this case, a sample from a sediment core).

In RStudio's "Environment" tab (usually at the top right of the screen), you should see a variable called `halifax_geochem` in the list. You can inspect it by clicking on the variable name, after which a tab will appear displaying the contents of the variable you just loaded. Clicking the little arrow to the left of the name will display the structure of the data frame, including the column names and some sample values. You can also do both of these things using the R commands `View()` and `glimpse()`, respectively. Also useful is the `head()` function, which will display the first few rows of a data frame.

```{r, eval=FALSE}
View(halifax_geochem) # will display a graphic table browser
```

```{r}
glimpse(halifax_geochem) # will display a text summary of the object
```

```{r}
head(halifax_geochem)
```

## Selecting Columns

One way to subset `valley_climate` is to subset by column, for which we will use the `select()` function. For example, we may only be interested in the mean temperature information, represented by the columns `station_name`, `year`, `month`, and `mean_temp`.

```{r, eval=FALSE}
mean_temp_data <- select(valley_climate, station_name, year, month, mean_temp)
```

The first argument to the `select()` function is the original data frame (in this case, `valley_climate`), and the remaining arguments are the names of the columns to be selected. To select the `station_name`, `year`, `month`, `mean_temp` and `total_precip` columns, you would use the following R command:

```{r, eval=FALSE}
temp_precip_data <- select(valley_climate, station_name, year, month, mean_temp, total_precip)
```

### Excercises

- Use `View()`, `glimpse()`, and `head()` to preview the two data frames we just created. Do they have the columns you would expect?
- Use `select()` to select `station_name`, `year`, `month`, and all of the columns containing temperature values, and assign it to the variable `temp_data`.

```{r, include = FALSE, eval = FALSE}
# using View, str, and head
View(mean_temp_data); str(mean_temp_data); head(mean_temp_data)
View(temp_precip_data); str(temp_precip_data); head(temp_precip_data)

# selecting all the temperature variables
temp_data <- select(valley_climate, mean_max_temp, mean_min_temp, mean_temp,
```

## Filtering Rows

Another way to subset `valley_climate` is by filtering rows using column values, similar to the filter feature in Microsoft Excel. This is done using the `filter()` function. For example, we may only be interested in July temperature for the two stations.

```{r, eval=FALSE}
july_data <- filter(valley_climate, month == 7)
```

Just like `select()`, the first argument to `filter()` is the original data frame, and the subsequent arguments are the conditions that each row must satisfy in order to be included in the output. Column values are referred to by the column name (in the above example, `month`), so to include all rows where the value in the `month` column is 7, we use `month == 7`. Passing multiple conditions means each row must satisfy all of the conditions, such that to obtain the data for July of 2000, we can use the following call to `filter()`:

```{r, eval=FALSE}
july_2000_data <- filter(valley_climate, month == 7, year == 2000)
```

It is very important that there are two equals signs within `filter()`! The `==` operator tests for equality (e.g. `(2 + 2) == 4`), whereas the `=` operator assigns a value or passes a named argument to a function, which is not what you're trying to do within `filter()`. Other common operators that are useful within filter are `!=` (not equal to), `>` (greater than), `<` (less than), `>=` (greater than or equal to), `<=` (less than or equal to), and `%in%` (tests if the value is one of several values). Using these, we could find out which observations had mean temperatures that were below freezing:

```{r, eval=FALSE}
freezing_observations <- filter(valley_climate, mean_temp < 0)
```

We could also find which observations occurred during the summer months (May, June, July, or August):

```{r, eval=FALSE}
summer_data <- filter(valley_climate, month %in% c(5, 6, 7, 8))
```

### Exercises

- Use `View()`, `str()`, and `head()` to preview the data frames we just created. Do they have the rows you would expect?
- Use `filter()` to find observations from the month of December where the mean temperature was above freezing.
- Are there any observations from the month of January where the mean temperature was below freezing?
- Filter `valley_climate` to include only observations from the months of December, January, February, and March and assign it to a variable name of your choosing.

```{r, include = FALSE, eval = FALSE}
# View, str, and head as above
filter(valley_climate, month == 12, mean_temp > 0)
filter(valley_climate, month == 1, mean_temp > 0)
winter_data <- filter(valley_climate, month %in% c(12, 1, 2, 3))
```

## Selecting and Filtering

Often we need to use both `select()` and `filter()` to obtain the desired subset of a data frame. To do this, we need to pass the result of `select()` to `filter()`, or the result of `filter()` to select. For example, we could create a data frame of mean temperature observations from the month of July one of two ways (you'll recall that we selected temperature columns in the data frame `mean_temp_data`, and we filtered for the month of July in the data frame `july_data`):

```{r, eval=FALSE}
july_temp <- filter(mean_temp_data, month == 7)
july_temp2 <- select(july_data, station_name, year, month, mean_temp)
```

### Exercises

- Inspect `july_temp` and `july_temp2` using `View()`, `str()`, and `head()`. Are they identical?
- Create a data frame of July total precipitation data and give it a variable name of your choosing. Do this by using `select()` followed by `filter()`, then using `filter()` followed by `select()`. Inspect the output to ensure the data frames are identical.

```{r, include = FALSE, eval = FALSE}
# the data frames july_temp and july_temp2 are identical
july_precip <- select(july_data, station_name, year, month, total_precip)
precip_data <- select(valley_climate, station_name, year, month, total_precip)
july_precip2 <- filter(precip_data, month == 7)
```

## The Pipe (%>%)

There is an easier way! Instead of creating intermediary variables every time we want to subset a data frame using `select()` and `filter()`, we can use the pipe operator (`%>%`) to pass the result of one function call to another. Thus, creating our `july_temp` data frame from above becomes one line with one variable assignment instead of two.

```{r, eval=FALSE}
july_temp3 <- valley_climate %>% 
  filter(month == 7) %>% 
  select(station_name, year, month, mean_temp)
```

What `%>%` does is pass the left side into the first argument of the function call on the right side. Thus, `filter(valley_climate, month == 7)` becomes `valley_climate %>% filter(month ==7)`. When using the tidyverse family of packages, you should use the pipe as often as possible! It usually makes for more readable, less error-prone code, and reduces the number of temporary variables you create that clutter up your workspace. When using `filter()` and `select()` with other tidyverse manipulations like `arrange()`, `group_by()`, `summarise()`, and `mutate()`, the pipe becomes indispensable.

### Exercises

- Inspect `july_temp3` to ensure it is identical to `july_temp`.
- Create a data frame of July total precipitation data using `valley_climate`, `filter()`, `select()`, and `%>%`. Is it identical to the data frame you created in the exercise above?

```{r, include = FALSE, eval = FALSE}
# the data frames july_temp and july_temp3 are identical
july_precip3 <- valley_climate %>% 
  filter(month == 7) %>% 
  select(station_name, year, month, total_precip)
```

## Sorting A Data Frame

Sometimes it is desirable to view rows in a particular order, which can be used to quickly determine min and max values of various parameters. You can do this in the interactive editor using `View()`, but sometimes rows need to be in particular order for plotting or other analysis. This is done using the `arrange()` function. For example, it may make sense to view `valley_climate` in ascending `year` and `month` order (most recent last):

```{r, eval=FALSE}
valley_climate %>%
  arrange(station_name, year, month)
```

Or descending year and month order (most recent first):

```{r, eval=FALSE}
valley_climate %>%
  arrange(station_name, desc(year), desc(month))
```

The `arrange()` function takes columns as arguments, surrounded by `desc()` if that column should be sorted in descending order.

## Distinct Values

It is often useful to know which values exist in a data frame. For example, I've told you that the two locations are for Kentville and Greenwood, but what are they actually called in the dataset? To do this, we can use the `distinct()` function.

```{r, eval=FALSE}
valley_climate %>%
  distinct(station_name)
```

The `distinct()` function can also take multiple columns, which will reduce the data frame to the unique combinations of those variables. For example, we could find out which years are represented in the data for each station.

```{r, eval=FALSE}
valley_climate %>%
  distinct(station_name, year)
```

The `distinct()` function can take any number of column names as arguments.

## Summarising A Data Frame

So far we have looked at subsets of `valley_climate`, but what if we want yearly averages instead of monthly averages? Using the tidyverse, we can `group_by()` the `station_name` and `year` column, and `summarise()`:

```{r, eval=FALSE}
valley_climate %>%
  group_by(station_name, year) %>%
  summarise(mean_temp = mean(mean_temp))
```

Here `group_by()` gets a list of columns, for which each unique combination of values will get one row in the output. `summarise()` gets a list of expressions that are evaluated for every unique combination of values defined by `group_by()` (e.g., `mean_temp` is the `mean()` of the `mean_temp` column for each station, for each year). Often, we want to include a number of summary columns in the output, which we can do by pasing more expressions to `summarise()`:

```{r, eval=FALSE}
valley_climate %>%
  group_by(station_name, year) %>%
  summarise(mean_temp = mean(mean_temp), 
            max_temp = max(extr_max_temp),
            min_temp = min(extr_min_temp))
```

You will notice that in 2002 the mean temperature for Kentville appears to be `NA`, or missing. This is because R propogates missing values unless you explicitly tell it not to. To fix this, you could replace `mean(mean_temp)` with `mean(mean_temp, na.rm = TRUE)`. Other useful functions to use inside `summarise()` include `mean()`, `median()`, `sd()`, `sum()`, `min()`, and `max()`. These all take a vector of values and produce a single aggregate value suitable for use in `summarise()`. One special function, `n()`, you can use (with no arguments) inside `summarise()` to tell you how many observations were aggregated to produce the values in that row.

```{r, eval=FALSE}
valley_climate %>%
  group_by(station_name, year) %>%
  summarise(mean_temp = mean(mean_temp, na.rm = TRUE), 
            max_temp = max(extr_max_temp, na.rm = TRUE),
            min_temp = min(extr_min_temp, na.rm = TRUE), 
            n = n())
```

Unsurprisingly, there are twelve observations per year, since there are twelve rows (one per month) in our original data frame. It's always a good idea to include `n()` inside `summarise()`, if nothing else as a check to make sure you've used `group_by()` with the correct columns.

### Excercises

- Assign the data frame we just created to a variable, and inspect it using `View()` and `str()`. Which years were the warmest? Which years were the coldest?
- Create a similar data frame to the one we just created but with precipitation. In which years did the most precipitation fall? (Hint: you will need to use the `sum()` function)
- Most of the data are from Greenwood with fewer years from Kentville. How many years are available for each station, and what is the range of years where data is available for each station? (Hint: you will need to `group_by(station_name)`)

## Extracting Columns

When we use `select()`, we get back a data frame, however occasionally we need one or a few of the vectors that make up the data frame (recall from the last tutorial that data frames are a collection of column vectors). If we needed *just* the temperature values, we can use the `$` operator to extract a column vector.

```{r}
halifax_geochem$C_percent
```

The problem with doing this is that our mean temperature values no longer have any context: half of them are from one station and half of them are from another, but this is not reflected without the other columns! Nevertheless, many R functions outside of the tidyverse require input as vectors (including many you've used so far, including `mean()`, `max()`, `min()`, etc.), and you will often see the `$` used in code written in other places to refer to columns. Functions in the tidyverse allow you to refer to columns by name (without the `$`) when used within specific functions (`summarise()` is a good example), so you should do this whenever you can! In the statistics tutorial we will start to use functions that are inconvenient to apply within the context of tidyverse and the pipe, but until then, get used to using `summarise()` and other tidyverse methods to call functions like `mean()`, `max()`, and `min()`. Later, we will introduce ways to calculate new columns using `mutate()`, which also let us refer to columns by name.

## Base R Subsetting vs. select() and filter()

In the wild, there are many ways to select columns and filter rows. I highly reccomend using `filter()` and `select()` to do this when writing new code, but you may see R code that subsets a data frame using square brackets in the form `my_data_frame[c("column_name_1", "column_name_2")]` or `my_data_frame[my_data_frame$column_name_1 > some_number, c("column_name_1", "column_name_2")]`. The latter is equivalent to `my_data_frame %>% select(column_name_1, column_name_2) %>% filter(column_name_1 > some_number)`. The tidyverse method of subsetting I find to be much more clear and far less error-prone, but it's worth knowing the other form so you can read R code written by others!

## Summary

In this tutorial we introduced the use of `select()`, `filter()`, `arrange()`, `distinct()`, and the pipe (`%>%`). We also used `group_by()` and `summarise()` to provide summary statistics from a data frame. These functions are the building blocks of other powerful tools in the tidyverse. For more information, see the [Data transformation](http://r4ds.had.co.nz/transform.html) chapter in [R for Data Science](http://r4ds.had.co.nz/). Another good resource is the [tidyverse, visualization, and manipulation basics](https://www.rstudio.com/resources/webinars/tidyverse-visualization-and-manipulation-basics/) tutorial from Garrett Grolemund.
