
# Creating Visualizations using ggplot

Intro

For more information, see the [data visualization](http://r4ds.had.co.nz/data-visualisation.html) chapter in [R for Data Science](http://r4ds.had.co.nz/).

## Prerequisites

The prerequisite for this tutorial is the `tidyverse` package. If this package isn't installed, you'll have to install it using `install.packages()`.

```{r, eval = FALSE}
install.packages("tidyverse")
```

Load the packages when you're done! If there are errors, you may have not installed the above packages correctly!

```{r}
library(tidyverse)
```

Finally, you will need to obtain the example data. Copy/paste the statement below to load the sample data for this tutorial (you can also [download the dataset as a csv](data/valley_climate.csv)).

```{r, include = FALSE}
# read local version to build
valley_climate <- read_csv("data/valley_climate.csv", col_types = cols(.default = col_guess()))
attr(valley_climate, "spec") <- NULL
```

```{r, eval = FALSE}
valley_climate <- read_csv("http://paleolimbot.github.io/r4wrs/data/valley_climate.csv", 
                         col_types = cols(.default = col_guess()))
```

It's worth mentioning a little bit about what this data frame contains, since we'll be working with it for the rest of this tutorial. Each row contains a number of parameters that are available on a monthly basis from two Environment Canada climate stations (Kentville Agricultural Research Station and Greenwood Station, both located in the [Annapolis Valley of Nova Scotia](https://en.wikipedia.org/wiki/Annapolis_Valley)). The `station_name`, `year`, and `month` columns identify where and when the values were measured, and the rest of the columns contain the measured values. The only column names that are slightly cryptic are `extr_max_temp` and `extr_min_temp`, which are the extreme maximum and minimum temperatures measured in that month, respectively.

## Using ggplot

The Grammar of Graphics (the "gg" in "ggplot") is a way of describing a graphic that is derived from data, which in R is done using the `ggplot()` function and its many friends. Unlike other plotting functions, `ggplot()` builds graphics from the data up (rather than starting with a template of a graphic and working backward). We will start with an example:

```{r}
ggplot(data = valley_climate, 
       mapping = aes(x = date, y = total_precip, colour = station_name)) +
  geom_line()
```

- What the structure of the `ggplot()` call is

Steps for plotting:

- Envision how you want your plot to look (draw it on paper if you have to!)
- Setup the data (`select()`, `filter()`)
- Setup your mapping (`aes()`)
- Choose your geoms (`geom_*()`)
- Make it look pretty

## Aesthetics

Categorical/Grouping Variables get mapped to X, Y, Colour, Shape, Linetype

Continuous Variables get mapped to X, Y, Colour, Size

## Geometries

Different geometry...

```{r}
ggplot(data = valley_climate, 
       mapping = aes(x = date, y = total_precip, colour = station_name)) +
  geom_point()
```

Multiple geometries!

```{r}
ggplot(data = valley_climate, 
       mapping = aes(x = date, y = total_precip, colour = station_name)) +
  geom_point() +
  geom_line()
```

## Facets

A way to make multiple groups

## Make it look pretty

### Labels

`labs()` function

### Scales

`scale_*_discrete()`, `scale_*_continuous()`

## Summary

Tutorial summary

For more information, see the [data visualization](http://r4ds.had.co.nz/data-visualisation.html) chapter in [R for Data Science](http://r4ds.had.co.nz/).
