
# Creating Visualizations using ggplot

Intro

For more information, see the [data visualization](http://r4ds.had.co.nz/data-visualisation.html) chapter in [R for Data Science](http://r4ds.had.co.nz/).

## Prerequisites

The prerequisite for this tutorial is the `tidyverse` package. If this package isn't installed, you'll have to install it using `install.packages()`.

```{r, eval = FALSE}
install.packages("tidyverse")
```

Load the packages when you're done! If there are errors, you may have not installed the above packages correctly!

```{r}
library(tidyverse)
library(readxl)
```

Finally, you will need to obtain the example data. You will need to [download the dataset as an excel sheet](data/halifax_geochem.xlsx) and put it in the "data" folder of your current project.

```{r}
# read local version to build
halifax_geochem <- read_excel("data/halifax_geochem.xlsx")
```

It's worth mentioning a little bit about what this data frame contains, since we'll be working with it for the rest of this tutorial.

TODO: describe the halifax_geochem dataset

## Using ggplot

The Grammar of Graphics (the "gg" in "ggplot") is a way of describing a graphic that is derived from data, which in R is done using the `ggplot()` function and its many friends. Unlike other plotting functions, `ggplot()` builds graphics from the data up (rather than starting with a template of a graphic and working backward). We will start with an example:

- What the structure of the `ggplot()` call is

Steps for plotting:

- Envision how you want your plot to look (draw it on paper if you have to!)
- Setup the data (`select()`, `filter()`)
- Setup your mapping (`aes()`)
- Choose your geoms (`geom_*()`)
- Make it look pretty

## Aesthetics

Categorical/Grouping Variables get mapped to X, Y, Colour, Shape, Linetype

Continuous Variables get mapped to X, Y, Colour, Size

## Geometries

## Facets

A way to make multiple groups

## Make it look pretty

### Labels

`labs()` function

### Scales

`scale_*_discrete()`, `scale_*_continuous()`

## Summary

Tutorial summary

For more information, see the [data visualization](http://r4ds.had.co.nz/data-visualisation.html) chapter in [R for Data Science](http://r4ds.had.co.nz/).
