
# Preparing and Loading Data {#prepare-load}

Intro

Draws from [data transformation](http://r4ds.had.co.nz/transform.html) and [data import](http://r4ds.had.co.nz/data-import.html) from [R for Data Science](http://r4ds.had.co.nz/).

## Prerequisites

The prerequisites for this tutorial are `tidyverse`. If this package isn't installed, you'll have to install it using `install.packages()`.

```{r, eval = FALSE}
install.packages("tidyverse")
```

Load the package when you're done! If there are errors, you may have not installed the above packages correctly!

```{r}
library(tidyverse)
```

Finally, you will need to obtain the example data. In this tutorial, we will use [Lake Arnold](https://www.google.com/maps/place/Lake+Arnold/@44.1305679,-73.9831746,11714m/data=!3m1!1e3!4m5!3m4!1s0x4ccaddaff2a5e819:0x392dc3e7c14fb40f!8m2!3d44.1310666!4d-73.9440977) diatom counts [@whitehead89], obtained from the [Neotoma database](https://www.neotomadb.org/). You will need to download the [tidy CSV version of the data](data/lake_arnold_valve_counts_tidy.csv), and the [Excel version of the data](data/lake_arnold_valve_counts.xlsx).

## Preparing data

Data comes in a nearly infinite variety of formats and configurations. Occasionally, you are in charge of entering data yourself, in which case you can plan ahead on how your data looks.

## Loading data

## Cleaning data

## Summary

Tutorial summary

Draws from [data transformation](http://r4ds.had.co.nz/transform.html) and [data import](http://r4ds.had.co.nz/data-import.html) from [R for Data Science](http://r4ds.had.co.nz/).
