
# Preparing and Loading your data into R

Intro

Draws from [data transformation](http://r4ds.had.co.nz/transform.html) and [data import](http://r4ds.had.co.nz/data-import.html) from [R for Data Science](http://r4ds.had.co.nz/).

## Prerequisites

The prerequisites for this tutorial are `tidyverse`. If this package isn't installed, you'll have to install it using `install.packages()`.

```{r, eval = FALSE}
install.packages("tidyverse")
```

Load the package when you're done! If there are errors, you may have not installed the above packages correctly!

```{r}
library(tidyverse)
```

Finally, you will need to obtain the example data. You will need to [download the dataset as an excel sheet](data/halifax_geochem.xlsx) and put it in the "data" folder of your current project.

```{r}
# read local version to build
halifax_geochem <- read_excel("data/halifax_geochem.xlsx")
```

It's worth mentioning a little bit about what this data frame contains, since we'll be working with it for the rest of this tutorial.

TODO: describe the halifax_geochem dataset

## Preparing data

Data comes in a nearly infinite variety of formats and configurations. Occasionally, you are in charge of entering data yourself, in which case you can plan ahead on how your data looks.

## Loading data

## Cleaning data

## Summary

Tutorial summary

Draws from [data transformation](http://r4ds.had.co.nz/transform.html) and [data import](http://r4ds.had.co.nz/data-import.html) from [R for Data Science](http://r4ds.had.co.nz/).
