
# Preparing and Loading your data into R

Intro

Draws from [data transformation](http://r4ds.had.co.nz/transform.html) and [data import](http://r4ds.had.co.nz/data-import.html) from [R for Data Science](http://r4ds.had.co.nz/).

## Prerequisites

The prerequisites for this tutorial are `tidyverse`. If this package isn't installed, you'll have to install it using `install.packages()`.

```{r, eval = FALSE}
install.packages("tidyverse")
```

Load the package when you're done! If there are errors, you may have not installed the above packages correctly!

```{r}
library(tidyverse)
```

Finally, you will need to obtain the sample data. We will be using the same sample data as before (`valley_climate`), but instead you will be using data files that contain this data in various forms. You can download this data in [comma-separated format (CSV)](data/valley_climate.csv), [tab-separated format (TSV)](data/valley_climate.tsv), and [Microsoft Excel format](data/valley_climate.xlsx), as well as the raw files (like you would download from [Environment Canada](http://climate.weather.gc.ca/)) for [Kentville](data/valley_climate_kentville.csv) and [Greenwood](data/valley_climate_greenwood.csv). You'll need all six files to do the excercises in this tutorial.

## Preparing data

Data comes in a nearly infinite variety of formats and configurations. Occasionally, you are in charge of entering data yourself, in which case you can plan ahead on how your data looks.

## Loading data

## Cleaning data

## Summary

Tutorial summary

Draws from [data transformation](http://r4ds.had.co.nz/transform.html) and [data import](http://r4ds.had.co.nz/data-import.html) from [R for Data Science](http://r4ds.had.co.nz/).
