
# Creating Stratigraphic Diagrams {#strat-diagrams}

Creating beautiful stratigraphic diagrams is difficult in any program, and while it is not necessarily easier to create beautiful digrams in R the first time, R becomes useful when the data used to create the graph get updated, requiring you to re-make the plot. In the course of a thesis, one can re-make a figure many, many times with various modifications. Learning to create publication-ready stratigraphic plots in R is a front-end investment of time that becomes useful after the first few rounds of revisions. This tutorial is designed to give you the tools to make such diagrams with your own data.

## Prerequisites

The prerequisites for this tutorial are the **tidyverse** package, the **gridExtra** package, and the **analogue** package. If you haven't installed the **analogue** or **gridExtra** packages yet, you'll have to install them using `install.packages()`.

```{r, eval = FALSE}
install.packages("analogue")
install.packages("gridExtra")
```

Load the tidyverse when you're done! We will load the other packages when we use functions that require them below.

```{r}
library(tidyverse)
```

Finally, you will need to obtain the example data. In this tutorial, we will use the [Lake Arnold](https://www.google.com/maps/place/Lake+Arnold/@44.1305679,-73.9831746,11714m/data=!3m1!1e3!4m5!3m4!1s0x4ccaddaff2a5e819:0x392dc3e7c14fb40f!8m2!3d44.1310666!4d-73.9440977) diatom counts [tidy CSV version of the data](data/lake_arnold_valve_counts_tidy.csv) [@whitehead89], and the [Halifax geochemistry data](data/halifax_geochem.csv). If you have these files downloaded you can load them yourself (see Tutorial \@ref(prepare-load)), or you can copy/paste the following code to load the two datasets.

```{r, include=FALSE}
halifax_geochem <- read_csv(
  "data/halifax_geochem.csv",
  col_types = cols(.default = col_guess())
)

arnold_counts_csv <- read_csv(
  "data/lake_arnold_valve_counts_tidy.csv",
  col_types = cols(.default = col_guess())
)
```

```{r, eval=FALSE}
halifax_geochem <- read_csv(
  "http://paleolimbot.github.io/r4paleolim/data/halifax_geochem.csv",
  col_types = cols(.default = col_guess())
)

arnold_counts <- read_csv(
  "http://paleolimbot.github.io/r4paleolim/data/lake_arnold_valve_counts_tidy.csv",
  col_types = cols(.default = col_guess())
)
```

## Workflow

It's worth mentioning a bit about how one might go about integrating R into one's figure-creating workflow. I suggest creating a file (something like `create_figures.R`) that loads the data, creates the figures, and saves the figures, within an RStudio project that also contains the data files. Keeping the data needed to create the figures and the script used to create the figures close means that you can reproduce the figure if you need to change the script (or change the data!), and using an RStudio project means you can move the folder around your computer (or to somebody else's computer) and your scripts won't change. I usually have an RStudio project with a subdirectory for `data-raw` (the data from the instrument/tech/emailed from other lab), `data` (the user-modified version of the files in `data-raw` that are R-friendly), and `figures` (the R-generated figures).

## General stratigraphic diagrams

Stratigraphic diagrams are at heart, a set of plots that share a Y axis. The Y-axis represents time, which can be expressed as depth or as some unit of actual time (e.g., AD 2018, or 1200 years BP), and the X-axis is the value of each parameter. We will use the **ggplot2** package to create these graphics for non-species data, and the **analogue** package to create the graphics for species composition data.

### Data preparation

Getting your data in a form that is usable by the plotting function is most of the battle to creating a figure. In the case of `ggplot()`, we need the data to be in a "parameter-long" form, with one row for each measurement (each point on the graph). In Tutorial \@ref(prepare-load) we learned how to use `gather()` to turn a "parameter-wide" data frame into a "parameter-long" data frame. We are mostly going to focus on plotting the Pockwock Lake core, so we will also filter the data to include only one core for now.

```{r}
halifax_geochem_meas <- halifax_geochem %>%
  filter(core_id == "POC15-2") %>%
  gather(key = param, value = value, -core_id, -depth_cm, -age_ad)

halifax_geochem_meas
```

### Plotting a single core

Plotting a single core with `ggplot()` involves several steps:

- The initial `ggplot()` call, where we set which columns will get mapped to the `x` and `y` values for each layer.
- Two layers: `geom_path()` and `geom_point()` (we don't use `geom_line()` because it was written by non-paleolimnologists and connects the line in odd ways)
- Specify how data are divided between facets using `facet_wrap()`. We need the scales to be free on the X-axis because each parameter has a different range of values, however the Y-axis should be the same for all facets.
- Reverse the Y-axis, so that a depth of 0 is at the top.
- Remove the X label, and set the Y label to "Depth (cm)".

```{r, warning=FALSE}
halifax_geochem_meas %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  facet_wrap(~param, scales = "free_x") +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)")
```

### Changing the colours

The default colour scheme in **ggplot2** is great for the web, but not so much for publication. I usually use `theme_bw()`, which is the black-and-white version of the default theme. There are several others, including `theme_minimal()`, `theme_classic()`, `theme_linedraw()`, and many more. You can add them to the end of your plot like so:

```{r, warning=FALSE}
halifax_geochem_meas %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  facet_wrap(~param, scales = "free_x") +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)") +
  theme_bw()
```

Alternatively, you can use the following line to set your theme for the rest of your R session! I usually set the theme to `theme_bw()` right under my call to `library(tidyverse)` in a script.

```{r}
theme_set(theme_bw())
```

### Facet configuration

The `facet_wrap()` function has `nrow` and `ncol` arguments that allow customization of how facets are laid out. Sometimes changing the figure size is also helpful, which can be done when the figure is exported using `ggsave()`. In general, you should only set one of `nrow` or `ncol`.

```{r, warning=FALSE}
halifax_geochem_meas %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  facet_wrap(~param, scales = "free_x", ncol = 4) +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)")
```

### Reordering facets

To control the orderering of the facets, we need to turn the `param` column into a `factor()`, which is kind of like a "multiple choice" vector that stores the order of its choices. We first need to create a character vector that defines the order, use `mutate()` to create a new column containing the facet label as a factor, then make `facet_wrap()` use that column to create the facets instead of `param`.

```{r, warning=FALSE}
facet_order <- c(
  "Ti_percent", "K_percent", 
  "C_percent", "C/N", 
  "d13C_permille", "d15N_permille"
)

halifax_geochem_meas %>%
  mutate(facet = factor(param, levels = facet_order)) %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  facet_wrap(~facet, scales = "free_x") +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)")
```

### Relabeling facets

When using `facet_wrap()`, **ggplot2** uses the value of each unique value in a column to create the label. This means that we need to rename each item to a suitable label prior to feeding it in to `ggplot()`. I think the easiest way to do this is `fct_recode()`, because it also keeps the order of the input (if you've already set the input to be a `factor()`), and you only have to rename the values that need renaming. (Note: I've use the unicode superscript 1-9 characters to get the superscript effect, because the other way is much harder. You can copy and paste these, or google [superscript 1 unicode](https://www.google.com/search?q=superscript+1+unicode)).

```{r, warning=FALSE}
halifax_geochem_meas %>%
  mutate(facet_label = fct_recode(
    param,
    "Ti (%)" = "Ti_percent",
    "K (%)" = "K_percent",
    "C (%)" = "C_percent",
    "δ¹³C (‰)" = "d13C_permille",
    "δ¹⁵N (‰)" = "d15N_permille"
  )) %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  facet_wrap(~facet_label, scales = "free_x") +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)")
```

If the unicode superscript effect doesn't work on your computer (it doesn't work for $\delta^{15}\text{N}$) on my computer), you will have to resort to `plotmath`. This is a custom way R invented to render complex formatting, and you can use it with **ggplot2** in various ways. The best way to implement this to start is to copy and paste, and check out `?plotmath` if you're curious what the syntax is. The key is to relabel the facets to very specific looking text (if in doubt, surround everything in single quotes like this: `'C/N'`), and then use `labeller = label_parsed` within `facet_wrap()` (also works in `facet_grid()`).

```{r, warning=FALSE}
halifax_geochem_meas %>%
  mutate(facet_label = fct_recode(
    param,
    "'Ti (%)'" = "Ti_percent",
    "'K (%)'" = "K_percent",
    "'C (%)'" = "C_percent",
    "delta ^ 13 * C" = "d13C_permille",
    "delta ^ 15 * N" = "d15N_permille",
    "'C/N'" = "C/N"
  )) %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  facet_wrap(~facet_label, scales = "free_x", labeller = label_parsed) +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)")
```

### Adding annotations

In general, adding text to the plot is difficult and not advisable. Adding horizontal lines and rectangles to highlight a particular region of the plot is much easier, and I have rarely had to resort to adding actual text to a plot using **ggplot2**. If this is critical to your application, you can use `ggsave()` to export a `.svg` file or `.pdf` file, and import it into your favourite vector drawing program (I happen to like [Inkscape](https://inkscape.org/en/)).

Horizontal lines can be added using `geom_hline()`. When using the `yintercept` argument, a horizontal line will be drawn on all panels.

```{r, warning=FALSE}
halifax_geochem_meas %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_hline(yintercept = c(7, 14, 21), alpha = 0.7, lty = 2) +
  geom_path() +
  geom_point() +
  facet_wrap(~param, scales = "free_x") +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)")
```

To only draw on some panels, you will need to create a tibble with a column that specifies the facet. A small example:

```{r, warning=FALSE}
hline_data <- tibble(
  param = "C/N",
  depth = 14
)

halifax_geochem_meas %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_hline(
    aes(yintercept = depth), data = hline_data,
    lty = 2, alpha = 0.7
  ) +
  geom_path() +
  geom_point() +
  facet_wrap(~param, scales = "free_x") +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)")
```

Rectangles can also be useful to draw in the background to highlight a range of depths. Rectangles also require tibble that specifies their appearance, which can contain the facet variable if the rectangle should only be drawn on some facets. Using `xmin = -Inf` and `xmax = Inf` ensures that the rectangles touch the edge of each facet.

```{r, warning=FALSE}
rect_data <- tibble(
  max_depth = 21,
  min_depth = 14
)

halifax_geochem_meas %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_rect(
    aes(ymin = min_depth, ymax = max_depth), data = rect_data,
    alpha = 0.3, xmin = -Inf, xmax = Inf, inherit.aes = FALSE
  ) +
  geom_path() +
  geom_point() +
  facet_wrap(~param, scales = "free_x") +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)")
```

### A second axis for ages

Depending on the application, ages can be used directly on the Y-axis, or you can include depth on the Y-axis and use a second axis on the right for ages. This can be done using `scale_y_reverse()` with the `sec.axis` argument. This axis is created using a one-to-one transformation, which requires the age-depth information as a tibble. Here I use a transformation function based on `approx()`, which, if you make sure your age-depth data frame is the same as the pockwock one, you should be able to safely copy and paste to add to your plot.

```{r, warning=FALSE}
pockwock_age_depth <- halifax_geochem %>%
  filter(core_id == "POC15-2") %>%
  select(depth_cm, age_ad)

halifax_geochem_meas %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  facet_wrap(~param, scales = "free_x") +
  scale_y_reverse(
    sec.axis = sec_axis(
      trans = ~approx(pockwock_age_depth, xout = .)$y,
      name = "Age (Year AD)",
      breaks = c(2000, 1950, 1900, 1850, 1800, 1750)
    ),
    expand = c(0, 0)
  ) +
  labs(x = NULL, y = "Depth (cm)")
```

### Multiple cores

Plotting multiple cores can happen in a few ways. First, you can map the `core_id` column to the colour aesthetic (you could use the linetype aesthetic if you need black and white output). You can specify the legend label using the `labs()` function.

```{r, warning=FALSE}
halifax_geochem %>%
  filter(core_id %in% c("POC15-2", "MAJ15-1")) %>%
  gather(key = param, value = value, -core_id, -depth_cm, -age_ad) %>%
  # add colour = core_id to aesthetics
  ggplot(aes(y = depth_cm, x = value, colour = core_id)) +
  geom_path() +
  geom_point() +
  facet_wrap(~param, scales = "free_x") +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)", colour = "Core ID")
```

Second, you can use `facet_grid()` instead of `facet_wrap()`. Here you can use age or depth on the y axis, although you can't use the second axis trick from above to have dates as a second axis.

```{r, warning=FALSE}
halifax_geochem %>%
  filter(core_id %in% c("POC15-2", "MAJ15-1")) %>%
  gather(key = param, value = value, -core_id, -depth_cm, -age_ad) %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  # use facet_grid instead of facet_wrap
  facet_grid(core_id ~ param, scales = "free_x") +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)", colour = "Core ID")
```

For both of these, you can use the same tricks we used to change the order of the facets to change the order and labelling of the end result.

### Overlapping x-axis labels

The default placement of labels in **ggplot2** can result in labels that overlap and look ugly in the final product. There are a few solutions, including (1) Rotate the labels by 90 degrees (this works in most circumstances):

```{r, warning=FALSE}
halifax_geochem %>%
  filter(core_id %in% c("POC15-2", "MAJ15-1")) %>%
  gather(key = param, value = value, -core_id, -depth_cm, -age_ad) %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  facet_grid(core_id~param, scales = "free_x") +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)", colour = "Core ID") +
  # modify axis labels
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5)
  )
```

You can also (2) make the font size for the labels smaller (the value is in points):

```{r, warning=FALSE}
halifax_geochem %>%
  filter(core_id %in% c("POC15-2", "MAJ15-1")) %>%
  gather(key = param, value = value, -core_id, -depth_cm, -age_ad) %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  facet_grid(core_id~param, scales = "free_x") +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)", colour = "Core ID") +
  # modify axis labels
  theme(
    axis.text.x = element_text(size = 7)
  )
```

Finally, you can (3) suggest a number of breaks to use in `scale_x_continuous()`:

```{r, warning=FALSE}
halifax_geochem %>%
  filter(core_id %in% c("POC15-2", "MAJ15-1")) %>%
  gather(key = param, value = value, -core_id, -depth_cm, -age_ad) %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  facet_grid(core_id~param, scales = "free_x") +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)", colour = "Core ID") +
  # x axis scale
  scale_x_continuous(breaks = scales::extended_breaks(n = 3))
```

### Saving a plot

You can use the same strategy we used in Tutorial \@ref(creating-visualizations-using-ggplot) to save the plots we created in this section. The easiest way to do this is to save the plot to an object, then use `ggsave()` to write the object to disk.

```{r}
halifax_geochem_plot <- halifax_geochem_meas %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  facet_wrap(~param, scales = "free_x") +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)")

ggsave(
  plot = halifax_geochem_plot, 
  filename = "geochem_plot_file_name.png",
  height = 4, width = 6.5,
  units = "in",
  dpi = 300
)
```

```{r, include=FALSE}
unlink("geochem_plot_file_name.png")
```

You can also save files as .pdf and .svg, but .png is probably the most reliable for importing into Word documents. If you need to edit the file in a vector drawing program afterward (or you need to give the final result to a journal), using .pdf or .svg is probably best. Note that exporting to .svg requires the **svglite** package.

## Exercises

- Make a simple stratigraphic diagram of the "BEN15-2" core. Use `age_ad` on the Y-axis. Your plot should look like this:

```{r, warning = FALSE}
halifax_geochem %>%
  filter(core_id == "BEN15-2") %>%
  gather(key = param, value = value, -core_id, -depth_cm, -age_ad) %>%
  ggplot(aes(y = age_ad, x = value)) +
  geom_path() +
  geom_point() +
  facet_wrap(~param, scales = "free_x") +
  labs(x = NULL, y = "Depth (cm)")
```

- Make a stratigraphic diagram of the "FCL16-1" and "LEM16-1" cores. For bonus points, use proper sub/superscripts for the delta parameters and keep C/N, C, and stable isotopes together, include units in parentheses for all parameters, and use `age_ad` on the Y-axis. In the end, your plot (with bonus marks) should look like this:

```{r echo = FALSE, warning = FALSE}
halifax_geochem %>%
  filter(core_id %in% c("FCL16-1", "LEM16-1")) %>%
  gather(key = param, value = value, -core_id, -depth_cm, -age_ad) %>%
  mutate(facet_label = fct_recode(
    param,
    "'Ti (%)'" = "Ti_percent",
    "'K (%)'" = "K_percent",
    "'C (%)'" = "C_percent",
    "delta ^ 13 * C" = "d13C_permille",
    "delta ^ 15 * N" = "d15N_permille",
    "'C/N'" = "C/N"
  )) %>%
  ggplot(aes(y = age_ad, x = value)) +
  geom_path() +
  geom_point() +
  facet_grid(core_id ~ facet_label, scales = "free_x", labeller = label_parsed) +
  labs(x = NULL, y = "Depth (cm)")
```

## Species composition diagrams

Species composition diagrams are like other types of strat diagrams, except they have long variable names and generally require scaling of the space such that 10% on one facet is the same distance as 10% on another facet. There are two approaches for this: using **ggplot2** is more verbose but more flexible (you can use most of the tips/tricks above to add secondary axes, horizontal lines, etc.), and using the `Stratiplot()` function in the **analogue** package is less verbose but requires that you have your data in a very specific form. We will go over both approaches here, but if you can make **ggplot2** work for you, the plot looks much nicer.

### Using ggplot2

#### Data setup

Similar to non-speices data, **ggplot2** needs the data to be in "parameter-long" form. Usually the value plotted on the X-axis of species composition plots is a relative abundance (in percent), which we can calculate using a grouped `mutate()`.

```{r}
arnold_counts <- arnold_counts_csv %>%
  gather(-age_bp, -depth_cm, key = taxon, value = valve_count) %>%
  group_by(depth_cm) %>%
  mutate(relative_abundance_percent = valve_count / sum(valve_count) * 100) %>%
  ungroup()

arnold_counts
```

Because there are many taxa, we will restrict our plot to those with a maximum relative abundance of 10% or more. We need this as a vector so that we can `filter()` the `arnold_counts` data above and change the `taxon` column to a `factor()` so that it is plotted in a specific order. Generally this order is some preference gradient, but that isn't a part of this particular dataset. Instead, we will order from least abundant to most abundant.

```{r}
arnold_common_taxa <- arnold_counts %>%
  group_by(taxon) %>%
  summarise(max_rel_abund = max(relative_abundance_percent)) %>%
  filter(max_rel_abund >= 10) %>%
  arrange(max_rel_abund) %>%
  pull(taxon)

arnold_common_taxa
```

```{r}
arnold_counts_common <- arnold_counts %>%
  filter(taxon %in% arnold_common_taxa) %>%
  mutate(taxon = factor(taxon, levels = arnold_common_taxa)) %>%
  arrange(taxon)

arnold_counts_common
```

#### Plotting one core

Using **ggplot2** to plot species composition strat diagrams is similar to plotting geochemical data, with three important differences: the geometry is usually a horizontal bar plot, the facet labels are so long that they need to be rotated, and the horizontal distance in each facet must be equal between facets (so that 10% relative abundance is the same on the leftmost facet as the rightmost facet). We solve the first by replacing `geom_path()` and `geom_point()` with `geom_segment()`, and the third by replacing `facet_wrap()` with `facet_grid()` using `space = "free_x"` to keep the distance on each facet comparable in the x direction. Setting the breaks on the x-axis helps keep the axes less cluttered and maintains the idea of equal space visually.

```{r}
ggplot(
  arnold_counts_common, 
  aes(y = depth_cm, x = relative_abundance_percent)
) +
  # draw horizontal lines of the appropriate length for each depth
  geom_segment(aes(xend = 0, yend = depth_cm), lwd = 1) +
  # facet by taxon, keeping distance on the x axis comparable
  # between facets
  facet_grid(~taxon, scales = "free_x", space = "free_x") +
  # have the same breaks for all x axes
  scale_x_continuous(breaks = c(0, 10, 20, 30)) +
  # reverse the y axis for depth
  scale_y_reverse() +
  labs(x = "Relative Abundance (%)", y = "Depth (cm)")
```

The problem of rotated and partially overlapping facet labels is one that takes some code to deal with. The good news is, the code is the same regardless of which plot you are trying to create, so you can copy/paste it for any plot that needs rotated facet labels. We do this in two steps: rotate and align the facet labels within `theme()`, and then use what can only be described as voodoo to eliminate the default clipping used by **ggplot2**.

```{r}
species_plot_obj <- ggplot(
  arnold_counts_common, 
  aes(y = depth_cm, x = relative_abundance_percent)
) +
  # draw horizontal lines of the appropriate length for each depth
  geom_segment(aes(xend = 0, yend = depth_cm), lwd = 1) +
  # facet by taxon, keeping distance on the x axis comparable
  # between facets
  facet_grid(~taxon, scales = "free_x", space = "free_x") +
  # reverse the y axis for depth
  scale_y_reverse() +
  # make all facets use the same break values
  # (helps with cluttered breaks)
  scale_x_continuous(breaks = c(0, 10, 20, 30)) +
  # set the x and y labels
  labs(x = "Relative Abundance (%)", y = "Depth (cm)") +
  # customize the appearance
  theme(
    # rotate the facet labels
    strip.text.x = element_text(angle = 60, hjust = 0, vjust = 0), 
    # turn off the label background
    strip.background = element_blank()
  )

# voodoo that makes it so that facet labels can overlap
# https://stackoverflow.com/questions/49740215/ggplot-facet-grid-label-cut-off
species_plot_grob <- ggplotGrob(species_plot_obj)
for(i in which(grepl("strip-t", species_plot_grob$layout$name))){
  species_plot_grob$grobs[[i]]$layout$clip <- "off"
}

# needed to draw the modified plot_grob
grid::grid.draw(species_plot_grob)
```

#### Saving a plot

Because we have modified the plot after it is has been built, we need to use the base-R way of saving things. This is done using the `pdf()`, `png()` and `svg()` functions (and calling `dev.off()` afterward). Call `grid::grid.draw()` in the middle like this:

```{r, results='hide'}
png(
  "my_species_plot.png", 
  width = 6.5, 
  height = 4, 
  res = 300,
  units = "in"
)
grid::grid.draw(species_plot_grob)
dev.off()

pdf(
  "my_species_plot.pdf", 
  width = 6.5, 
  height = 4
)
grid::grid.draw(species_plot_grob)
dev.off()
```

```{r, include = FALSE}
unlink("my_species_plot.pdf")
unlink("my_species_plot.png")
```

#### Adjusting spacing

For more fine-tuned control of the appearance, you can use various arguments within `theme()`. The two that I've used in this situation are `plot.margin` to add some padding to the right (this helps with the occasional cutoff rightmost facet label), and `panel.spacing.x` to adjust the spacing between panels.

```{r, eval = FALSE}
  ... +
  theme(
    ...,
    # add some margin to the right of the plot so that the rightmost label
    # isn't cut off
    # c(top, right, bottom, left)
    plot.margin = unit(c(0, 0.5, 0, 0), "inches"),
    # adjust spacing between facets
    panel.spacing.x = unit(0.05, "inches")
  )
```

#### Non-species data

Often strat diagrams include some other kind of data in addition to the species data, like a PC1 summary or diatom-inferred pH. I'll calculate a few diversity indicies for this dataset as a demonstration.

```{r}
species_diversity <- arnold_counts %>%
  group_by(depth_cm) %>%
  summarise(
    richness = sum(valve_count != 0),
    simpsons = 1 - sum((relative_abundance_percent / 100) ^ 2),
    hill_n2 = 1 / sum((relative_abundance_percent / 100) ^ 2)
  ) %>%
  gather(key = param, value = value, -depth_cm)

diversity_plot <- ggplot(species_diversity, aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  facet_wrap(~param, scales = "free_x") +
  scale_y_reverse() +
  labs(x = NULL, y = "Depth (cm)")

diversity_plot
```

When using **ggplot2**, there is no easy way to combine this with a species plot (you will see a way to do this in the **analogue** package [below](#strat-species-analogue)). The only way I know of to do this is to with **ggplot2** is to (1) modify the non-species data such that it has no y-axis label, no y-axis ticks, and no y-axis break labels, (2) remove the plot background from the second plot (to avoid clipping), (3) fudge the plot margins on the second plot so that the plots align properly (below I set the top and bottom margins to 1.74 and 0.30 inches, respectively, but you will have to fudge this yourself), and (4) use the `grid.arrange()` function from the **gridExtra** package to combine the two plots (you can set the relative widths, but you should have the number of rows be 1). Make sure to use the "grob" version of the species plot, which kept the species names from being clipped. Make sure that your depth values actually align before removing the axis labels completely!

```{r, message=FALSE}
gridExtra::grid.arrange(
  species_plot_grob,
  diversity_plot + 
    labs(y = NULL) +
    scale_y_reverse(labels = NULL) +
    theme(
      plot.margin = unit(c(1.74, 0, 0.30, 0), "inches"),
      axis.ticks.y = element_blank(),
      plot.background = element_blank()
    ),
  nrow = 1,
  widths = c(2.5, 1)
)
```

You can save this plot by enclosing the `grid.arrange()` call in `png()` (or `pdf()`) and `dev.off()`.

### Using analogue::Stratiplot {#strat-species-analogue}

The [**analogue** package](https://github.com/gavinsimpson/analogue) by [Gavin Simpson](https://www.fromthebottomoftheheap.net/) has a number of useful functions for paleolimnologists, including a `Stratiplot()` function designed to produce stratigraphic plots. I find the **ggplot2** version to be more flexible, but the default look and feel works for you, `Stratiplot()` is definitely easier. Because we are using functions from another package, we have to load the package using `library()`.

```{r}
# install using install.packages("analogue")
library(analogue)
```

#### Data preparation

The data required by `Stratiplot()` is a data frame of relative abundances (as columns). To calculate this from our filtered data, we can use the `spread()` function to get the data in "parameter wide" form.

```{r}
arnold_rel_abundance_wide <- arnold_counts_common %>%
  select(depth_cm, age_bp, taxon, relative_abundance_percent) %>%
  spread(key = taxon, value = relative_abundance_percent)

arnold_rel_abundance_wide
```

#### Plotting a single core

The `Stratiplot()` function takes two arguments: the first is a version of the data frame we just created without any of the depth information. The second is the depth information as a vector.

```{r}
Stratiplot(
  arnold_rel_abundance_wide %>% select(-depth_cm, -age_bp),
  arnold_rel_abundance_wide$depth_cm
)
```

There are many arguments to `Stratiplot()`, but I found for this example that the following values worked for this particular dataset. In particular, increasing the `topPad` parameter was needed to keep the species labels from being cutoff.

```{r, message=FALSE, warning=FALSE}
Stratiplot(
  arnold_rel_abundance_wide %>% select(-depth_cm, -age_bp),
  arnold_rel_abundance_wide$depth_cm, 
  # sets the x and y labels
  ylab = "Depth (cm)", 
  xlab = "Relative abundance (%)",
  # adds padding to the top of the plot
  # to fix cut-off taxa names
  topPad = 10, 
  # make the plot type a "bar" plot
  type = "h", 
  # make the bar colour black
  col = "black"
)
```

#### Non-species data

Including non-species data is similar to including species data using `Stratiplot()`, but the `varTypes` argument needs to specify that the non-species variables should have independently sized axes. This should be a vector with the same number of elements as variables in the plot (I've use `rep()` to repeat "relative" and "absolute" the correct number of times.)

```{r}
species_diversity <- arnold_counts %>%
  group_by(depth_cm) %>%
  summarise(
    richness = sum(valve_count != 0),
    simpsons = 1 - sum((relative_abundance_percent / 100) ^ 2),
    hill_n2 = 1 / sum((relative_abundance_percent / 100) ^ 2)
  )

arnold_rel_abundance_wide_div <- cbind(
  arnold_rel_abundance_wide,
  species_diversity %>% select(-depth_cm)
)

Stratiplot(
  arnold_rel_abundance_wide_div %>% select(-depth_cm, -age_bp),
  arnold_rel_abundance_wide_div$depth_cm, 
  varTypes = c(rep("relative", 11), rep("absolute", 3)),
  ylab = "Depth (cm)", 
  xlab = "Relative abundance (%)",
  topPad = 10, 
  type = "h", 
  col = "black"
)
```

#### Saving a plot

You can save these plots by enclosing the `Stratiplot()` call in `png()` (or `pdf()`) and `dev.off()`:

```{r, results='hide'}
png(
  "my_species_plot.png", 
  width = 6.5, 
  height = 4, 
  res = 300,
  units = "in"
)
Stratiplot(
  arnold_rel_abundance_wide %>% select(-depth_cm, -age_bp),
  arnold_rel_abundance_wide$depth_cm, 
  # sets the x and y labels
  ylab = "Depth (cm)", 
  xlab = "Relative abundance (%)",
  # adds padding to the top of the plot
  # to fix cut-off taxa names
  topPad = 10, 
  # make the plot type a "bar" plot
  type = "h", 
  # make the bar colour black
  col = "black"
)
dev.off()
```

## Exercises

- Use either **ggplot2** or **analogue** to plot the species `c("Surirella delicatissima", "Anomoeoneis serians var. brachysira", "Navicula subtilissima", "Fragilaria pinnata", "Fragilaria brevistriata")`. For bonus points, use **ggplot2** and rotate the axis labels.

```{r, echo=FALSE}
arnold_exercise_taxa <- c(
  "Surirella delicatissima", "Anomoeoneis serians var. brachysira", 
  "Navicula subtilissima", "Fragilaria pinnata", "Fragilaria brevistriata"
)

arnold_exercise_data <- arnold_counts %>%
  filter(taxon %in% arnold_exercise_taxa) %>%
  mutate(taxon = factor(taxon, levels = arnold_exercise_taxa))

exercise_plot_obj <- ggplot(
  arnold_exercise_data, 
  aes(y = depth_cm, x = relative_abundance_percent)
) +
  # draw horizontal lines of the appropriate length for each depth
  geom_segment(aes(xend = 0, yend = depth_cm), lwd = 1) +
  # facet by taxon, keeping distance on the x axis comparable
  # between facets
  facet_grid(~taxon, scales = "free_x", space = "free_x") +
  # reverse the y axis for depth
  scale_y_reverse() +
  # make all facets use the same break values
  # (helps with cluttered breaks)
  scale_x_continuous(breaks = c(0, 10, 20, 30)) +
  # set the x and y labels
  labs(x = "Relative Abundance (%)", y = "Depth (cm)") +
  # customize the appearance
  theme(
    # rotate the facet labels
    strip.text.x = element_text(angle = 60, hjust = 0, vjust = 0), 
    # turn off the label background
    strip.background = element_blank()
  )

# voodoo that makes it so that facet labels can overlap
exercise_plot_grob <- ggplotGrob(exercise_plot_obj)
for(i in which(grepl("strip-t", exercise_plot_grob$layout$name))){
  exercise_plot_grob$grobs[[i]]$layout$clip <- "off"
}

# needed to draw the modified plot_grob
grid::grid.newpage()
grid::grid.draw(exercise_plot_grob)

# stratiplot
arnold_exercise_wide <- arnold_exercise_data %>%
  select(depth_cm, age_bp, taxon, relative_abundance_percent) %>%
  spread(key = taxon, value = relative_abundance_percent)

Stratiplot(
  arnold_exercise_wide %>% select(-depth_cm, -age_bp),
  arnold_exercise_wide$depth_cm, 
  ylab = "Depth (cm)", 
  xlab = "Relative abundance (%)",
  topPad = 10, 
  type = "h", 
  col = "black"
)
```

```{r, include = FALSE}
unlink("my_species_plot.png")
```

## Summary

In this tutorial, we used **ggplot2** and the **analogue** package to create (and save) stratigraphic diagrams for geochemical and species composition data. Unforunately, I don't know of any good resources currently available for creating stratigraphic plots in R, although there is an [R session at IPA-IAL](https://ipa-ial.geo.su.se/courses-and-meetings) this year.
