
# Creating Strat diagraoms {#strat-diagrams}

Intro

## Prerequisites

The prerequisites for this tutorial are the **tidyverse** package, the **analogue** package, and the **rioja** package. If you haven't installed the **analogue** and **rioja** packages yet, you'll have to install them using `install.packages()`.

```{r, eval = FALSE}
install.packages(c("rioja", "analogue"))
```

Load the tidyverse when you're done! We will load the other packages when we use functions that require them below.

```{r}
library(tidyverse)
```

Finally, you will need to obtain the example data. In this tutorial, we will use the [Lake Arnold](https://www.google.com/maps/place/Lake+Arnold/@44.1305679,-73.9831746,11714m/data=!3m1!1e3!4m5!3m4!1s0x4ccaddaff2a5e819:0x392dc3e7c14fb40f!8m2!3d44.1310666!4d-73.9440977) diatom counts [tidy CSV version of the data](data/lake_arnold_valve_counts_tidy.csv) [@whitehead89], and the [Halifax geochemistry data](data/halifax_geochem.csv). If you have these files downloaded you can load them yourself (see Tutorial \@ref(prepare-load)), or you can copy/paste the following code to load the two datasets.

```{r, include=FALSE}
halifax_geochem <- read_csv(
  "data/halifax_geochem.csv",
  col_types = cols(.default = col_guess())
)

arnold_counts <- read_csv(
  "data/lake_arnold_valve_counts_tidy.csv",
  col_types = cols(.default = col_guess())
)
```

```{r, eval=FALSE}
halifax_geochem <- read_csv(
  "http://paleolimbot.github.io/r4paleolim/data/halifax_geochem.csv",
  col_types = cols(.default = col_guess())
)

arnold_counts <- read_csv(
  "http://paleolimbot.github.io/r4paleolim/data/lake_arnold_valve_counts_tidy.csv",
  col_types = cols(.default = col_guess())
)
```

## Data preparation

- TODO: How to rename columns so that they look pretty, how to use `gather()` and `spread()` to get data in the right form for the various functions.

## Using ggplot

### Single core

```{r}
halifax_geochem %>%
  filter(core_id == "POC15-2") %>%
  gather(-core_id, -depth_cm, -age_ad, key = param, value = value) %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  facet_wrap(~param, scales = "free_x") +
  scale_y_reverse()
```

### Multiple cores

- Use colour aesthetic, age on y axis

```{r}
halifax_geochem %>%
  filter(core_id %in% c("POC15-2", "MAJ15-1")) %>%
  gather(-core_id, -depth_cm, -age_ad, key = param, value = value) %>%
  ggplot(aes(y = age_ad, x = value, colour = core_id)) +
  geom_path() +
  geom_point() +
  facet_wrap(~param, scales = "free_x")
```

- Use facet_grid, age or depth on y-axis

```{r}
halifax_geochem %>%
  filter(core_id %in% c("POC15-2", "MAJ15-1")) %>%
  gather(-core_id, -depth_cm, -age_ad, key = param, value = value) %>%
  ggplot(aes(y = depth_cm, x = value)) +
  geom_path() +
  geom_point() +
  facet_grid(core_id~param, scales = "free_x") +
  scale_y_reverse()
```

### Species data

```{r}
arnold_counts_long <- arnold_counts %>%
  gather(-age_bp, -depth_cm, key = taxon, value = valve_count) %>%
  group_by(depth_cm) %>%
  mutate(relative_abundance = valve_count / sum(valve_count) * 100) %>%
  ungroup() %>%
  group_by(taxon) %>%
  filter(max(relative_abundance) > 20) %>%
  ungroup()
```

```{r}
p <- ggplot(arnold_counts_long, aes(y = depth_cm, x = relative_abundance)) +
  geom_segment(aes(xend = 0, yend = depth_cm)) +
  geom_path() +
  facet_grid(~taxon, scales = "free_x", space = "free_x") +
  scale_y_reverse() +
  scale_x_continuous(breaks = c(0, 5, 10, 15, 20, 25, 30)) +
  theme_minimal() +
  theme(
    strip.text.x = element_text(angle = 45, hjust = 0, vjust = 0)
  )

# gt <- ggplot_gtable(ggplot_build(p))
# gt$layout$clip[gt$layout$name=="panel-1-1"] = "off"
# grid::grid.draw(gt)
p
```

## Using analogue::stratiplot

## Using rioja::strat.plot

## Summary

Tutorial summary
